\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[colorlinks=true]{hyperref}

\begin{document}
\begin{center}
  {\bfseries Assignment \#4}\\
  Introduction to GR, 2020 Fall\\
  International Centre for Theoretical Sciences\\
  Instructor: Prof. Bala Iyer, Tutor: Md Arif Shaikh\footnote{\href{mailto: arif.shaikh@icts.res.in}{arif.shaikh@icts.res.in}}\\
  Due on November 1, 2020 11:59 PM.
\end{center}
\hrule

\begin{enumerate}
\item From d'Inverno's book, 10.3, 10.5, 10.6, 10.7, 10.8, 10.9, 10.10.
\item From d'Inverno's book, 12.2, 12.4, 12.5, 12.6, 12.9, 12.10, 12.11, 12.12.
\item From d'Inverno's book, 14.1, 14.2, 14.6, 14.9, 14.13, 14.14, 14.15.
\item
  \begin{enumerate}
  \item Consider a spacetime with a Killing vector $\xi$. Show that if
    the energy-momentum tensor $T^{\mu\nu}$ satisfies the conservation
    equation
    \begin{equation}
      \label{eq:conservation-of-Tmunu}
      \nabla_\mu T^{\mu\nu} = 0,
    \end{equation}
    then the four-vector $Q^\mu := T^{\mu\nu}\xi_\nu$ also satisfies a conservation law $\nabla_\mu Q^\mu = 0$.
  \item Now let us consider a flow consisting of perfect fluid and governed by adiabatic equation of state. Using first law of thermodynamics and specific enthalpy $h := (p + \varepsilon)/\rho$, where $p$ is pressure density, $\varepsilon$ is the internal energy density and $\rho$ is the fluid density, show that the conservation law in Eq. (\ref{eq:conservation-of-Tmunu}) provides
    \begin{equation}
      \label{eq:Euler-eq-enthalpy}
      u^\mu \nabla_\mu (h u_\nu) = \frac{\nabla_\nu p}{\rho}
    \end{equation}
  \item Using Eq. (\ref{eq:Euler-eq-enthalpy}), find the expression for $\pounds_u(hu_\mu)$. Contracting it with $\xi$, find the expression for $\pounds_u(hu_\mu \xi^\mu)$ in terms of $\pounds_\xi p$ and $\pounds_\xi h$.
  \item 
    Let's also assume that the flow is invariant under the same symmetry group of the spacetime which provieds the Killing vector $\xi$. Mathematically this gives the condition
    \begin{equation}
      \label{eq:symmetry-of-flow}
      \pounds_\xi(B) = 0,
    \end{equation}
    where B is any tensor field associated with the flow, e.g., pressure, density etc. Show that this implies that $h u_\mu \xi^\mu$ is conserved along the flow lines.
  \item Find the conserved quantity along the flow lines when the spacetime is stationary and the flow is stationary. This is the {\itshape general relativistic Bernoulli constant}.
  \item Find the Newtonian limit and check that the quantity reduces to the familiar expression of Bernoulli equation.
\end{enumerate}
  
\end{enumerate}

\end{document}